---
title: "Class 5: Data Viz with ggplot"
author: "Vivian Bach (PID:A18497911)"
format: pdf
toc: true
---

## Background

There are lot's of ways to make figures in R. These include so-called "base R" graphics (e.g `plot()`) and tones of add-on packages like **ggplot2**.

For example here we make the same plot with both:

```{r}
head(cars)
```

```{r}
plot(cars)
```
First I need to install the package with the commans 
`install.packages()`
install.packages("ggplot2")
library(ggplot2)

>**N.B.** We never run an install cmd in a quarto code chunk or we will end up re-installing packages many many time- which is not what we want! 

Everytime we want to use one of these "add-on" packages we need to load it up in R with the `library()` function: 
```{r}
library(ggplot2)
```

```{r}
ggplot(cars)
```




Every ggplot needs ar least 3 things:
-the **data**, the stuff you want plotted
-the **aes**thitics, hoe the data map to the plot
-the **geom**etry, the type of plot 
```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() 
```


Add a line to better show relationship between speed and dist 
```{r}
p <- ggplot(cars) + 
  aes(x=speed, y=dist) + 
  geom_point() + 
  geom_smooth(method="lm", se=FALSE) +
  labs(title="Stopping distance of old cars",
       subtitle = "Data from the `cars` object",
       x="Speed (mph)", 
       y="Distance (ft)")
```

```{r}
p + theme_bw()
```

## Gene expression plot 
we 
```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```
A first version 

```{r}
ggplot(genes) +
  aes(Condition1, Condition2) +
  geom_point()
```
```{r}
head(genes$State,3)
```

Version 2 let's color by `State` so we can see the up and down significant genes compared ti all the "unchanging genes

```{r}
ggplot(genes) +
  aes(Condition1, Condition2, col=State) +
  geom_point()
```
Version 3 plot, lets motify the default colors to something we like 
```{r}
ggplot(genes) +
  aes(Condition1, Condition2, col=State) +
  geom_point() + 
  scale_colour_manual(values =c("red",
                                "purple",
                                "pink")) + 
  labs(x="Control(no drug)",
       y="Drug",
       title = "Gene Expression Changes upon GLP-1 drug") + 
  theme_bw()
                                
```
```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder, 3)
```
```{r}
ggplot(gapminder) + 
  aes(x=gdpPercap, y=lifeExp, col=continent) +
  geom_point(alpha=0.3)
```
Let's "facet" (i.e make a seperate plot) by cotinent rather than the big hot mess above
```{r}
ggplot(gapminder) + 
  aes(x=gdpPercap, y=lifeExp, col=continent) +
  geom_point(alpha=0.3) +
  facet_wrap(~continent)
```

```{r}
nrow(gapminder)
```
I want to "filter" down to a subset of this data. I will use the **dplyr** package to help me. 

First I need to install it and then load it up...

```{r}
library(dplyr)

gapminder_2007 <- filter(gapminder, year==2007)
head(gapminder_2007)
```
```{r}
filter(gapminder_2007, country=="United States")
```
> Q. Make a plot comparing 1977 and 2007 for all countries

```{r}
input <- filter(gapminder, year %in% c(1977, 2007))
head(input)
```
```{r}
gapminder_1977 <- gapminder %>% filter(year==1977 | year==2007)

ggplot(gapminder_1977) + 
  geom_point(aes(x = gdpPercap, y = lifeExp, color=continent), alpha=0.7) + 
  scale_size_area(max_size = 10) +
  facet_wrap(~year)
```

